plyranges: a grammar for manipulating genomics data

### 2018-02-25 paper.Rmd
not happy with that idea ---
updates

---
title: "plyranges: a grammar for manipulating genomics data" 
author:
  - name: Stuart Lee
    email: stuart.lee1@monash.edu
    affiliation: Monash University
  - name: Michael Lawrence
    email: michael.lawrence@gene.com
    affiliation: Genentech
  - name: Di Cook
    email: di.cook@monash.edu
    affiliation: Monash University
address:
  - code: Monash University
    address: Department of Econometrics and Business Statistics, Clayton, Victoria, Australia
  - code: Genentech
    address: Bioinformatics and Computational Biology, Genentech, Inc., South San Francisco, California, United States of America 
abstract: |
    The Bioconductor project has created many powerful abstractions for reasoning about genomics data, such as the _Ranges_ data structures for representing genomic intervals. By recognising that these data structures follow 'tidy' data principles we have created a grammar of genomic data manipulation that defines verbs for performing actions on and between genomic interval data. This grammar simplifies performing common genomic data analysis tasks via method chaining, type consistency and results in creating human readable pipelines. We have implemented this grammar as an Bioconductor/R package called plyranges.

bibliography: references.bib
output: 
  rticles::plos_article:
csl: plos.csl
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = "#>", 
                      fig.width = 5, fig.height = 3, fig.align = "center")
options(digits = 3)
library(ggbio)
library(ggplot2)
library(plyranges)
```
# Introduction

Genomic data may be naturally represented as sets of pairs of integers 
corresponding to the start and end points of sequences. Further information
such as strandedness and chromosome name may be added to these sets to 
provide biological context. Because of the flexibility of this representation 
supplemental information such as measurements obtained from an experimental 
assay or annotations from a genome database can be joined to their relevant
sequences. In the Bioconductor/R packages `IRanges` and `GenomicRanges` these 
representations have been implemented as a suite of data structures called
_Ranges_ @Lawrence2013-wg. These data structures cover many common data types 
encountered in bioinformatics analyses - a gene can be represented with its 
coordinates, along with its identifier and the identifiers of its exons; or
an RNA-seq experiment may be represented as sets of genes with a matching count 
column.

The Bioconductor infrastructure for computing with genomic ranges are highly 
effecient and powerful, however the application programming interface (API) for 
performing analysis tasks with _Ranges_ is complex due to its large number of
methods and classes. It also makes resulting scripts written 
difficult for a non-programmer to read and reason about. An alternative approach 
would be to implement a domain specific language (DSL) as a fluent interface 
built on top _Ranges_. The goal of fluent interface is to enable users to write 
human-readable code via method chaining and consistent function returns. Fluent 
interfaces fit naturally in the context of Bioinformatics workflows because they 
enable writing succinct pipelines.

Several other DSLs have been implemented to reason about genomics data. Broadly, 
these are either implemented as query languages or as command line tools 
embedded in the unix environment ^[other ideas to mention GROK @Ovaska2013-gd]. 
An example of the former is the Genome Query Language (GQL) and its distributed 
implementation GenAp which use an SQL-like syntax for fast retrieval of 
information from genomic databases and BAM files @Kozanitis2014-va; @Kozanitis2016-bm. 
Another  example is the Genometric Query Language (GMQL) which implements a 
relational algebra for combining big genomic datasets @Kaitoua2017-pw.  
The command line application 
BEDtools develops an extensive algebra for performing arithmetic between two or 
more sets of genomic regions  @Quinlan2010-gc. It also has a python interface 
which simplifies constructing scripts for performing analyses based on BEDTools 
@Dale2011-js. 
<!--- probably should also mention something about their cons, and more detail --->

The abstraction provided by the _Ranges_ data structures aligns with the concept
of tidy data @Wickham2014-jc. The tidy data pattern is useful because it allows 
us to see how the data relates to the design of an experiment
and the variables measured. Consequently, it makes both the modelling and 
manipulation of data more systematic. The _Ranges_ data structure follows 
this abstraction: it is a rectanglular table corresponding to a single biological 
context. Each row contains a single observation and each column a variable about
that observation.

The tidy data abstraction has motivated the development of `plyranges`
a grammar of genomic data manipulation based on the _Ranges_ data structures. It
implements and extends the grammar defined by the R package `dplyr` @Wickham2017-dplyr. 
The grammar provides a consistent way of interacting with and analysing genomic 
data via methods for constructing, grouping, mutating, filtering, and 
summarising _Ranges_ and an algebra for reasoning about actions on _Ranges_
and relationships between _Ranges_.

@@

The _plyranges_ API implements a domain specific language using the existing
_IRanges_ and _GenomicRanges_ packages in Bioconductor as a backend. Consequently,
our API still has the speed and effeciency of the aforementioned packages but
with a more coherent interface. The API also
extends the grammar elements in _dplyr_ for performing genomic specific
manipulations such as finding overlapping regions or nearest neighbour regions
between many _Ranges_. The _plyranges_ API is specifically designed to enable
fast interactive analysis of _Ranges_ objects but can also be used for scripting 
genomic data workflows.

We have desgined the API to be fluent. Every function call corresponds to an
action on a _Ranges_ object (they are named verbs) and where possible
functions have few arguments. Each verb is constructed
to enable a tab completion based workflow. Both of these aspects reduce the 
cognitive load on a new user  since most manipulations can be performed with a 
vocabulary of several verbs, rather than having to memorise functions with many 
arguments that are nouns (as is required in the existing Bioconductor packages). 
This is also has the advantage of allowing users to write human readable code
because verbs describe what the code is doing rather than how its doing it.

Workflows can be composed by chaining verbs together via the pipe operator,`%>%` 
(exported from the R package _magrittr_). This is possible because every 
function call is endomorphic: when the input is _Ranges_ object the output
will also be a _Ranges_ object. One advantage of this static typing is that it 
does not  require any additional learning of classes beyond _Ranges_ and 
the _DataFrame_ classes. This is similar to the bedtools API, where the
output is usually a BED file. However, it strongly deviates from the design of 
the _Ranges_ Bioconductor packages, where many methods return a new class
upon return. The Bioconductor design enables effecient computing as users
are exposed to low-level features of its API which _plyranges_ trys to abstract
away. Method chaining via the pipe operator can also be difficult to debug,
as there multiple points of failure.

In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
use of non-standard evaluation in R via the _rlang_ package. Simply, this
means that compuations are performed and evaluated in the context of the 
Ranges objects, which emphasises the interactive nature of the API. This has
the disadvantage that programming with _plyranges_ becomes more difficult 
because a user needs to capture expressions inside function calls and then
unquote them.
<!--- need to make this clearer? --->

## Working with Ranges

### Construction and Import/Output

The _plyranges_ API provides the methods `as_granges()` and `as_iranges()`
for constructing Ranges from tabular data structures, such as the _data.frame_ in
base R. These methods use non-standard evaluation so columns in a _data.frame_
can be modified before a _Ranges_ object is created. The API also has
convienence methods for annotating or extracting annotations from Genomic Ranges
objects with the `set_genome_info()` and `get_genome_info()` functions.

There is a consistent framework for reading and writing 
files from and to common genomic data formats, using the `rtracklayer` package
as a backend. The methods are implemented in the `read_/write_` family of 
functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
narrowPeaks, GFF/GTF, WIG and BigWig files.

As an example, we use the Bioconductor package AnnotationHub to search for 
BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project.
We choose to focus on assays for primary T CD8+ memory cells from peripheral 
blood. We can then read the BigWig file  corresponding to the H3 lysine 27 
 trimethylation (H3K27Me3) 
methylation mark over chromosome 10. First we gather the BigWig file
and extract its annotation information and filter it to chromosome 10.

```{r load-bw, cache = TRUE}
# E048 corresponds to CD8+ T cells from peripheral blood
library(plyranges)
library(AnnotationHub)
q <-query(AnnotationHub(), c("EpigenomeRoadMap", "E048"))

bw_file <- q[["AH33458"]]

chr10_ranges <- bw_file %>% 
  get_genome_info() %>%
  filter(seqnames == "chr10")
```

Then we read the BigWig file only extracting scores if they overlap chromosome
10. We then add annotation information to ensure we know we are working
with the hg19 genome build.

### Manipulation of Ranges

The _plyranges_ API exports the six core verbs from the _dplyr_ package
and modifys them for use with simple _Ranges_ and _Genomic Ranges_. The
verb `mutate()` takes a Ranges object and a set of name-value pairs
and generates a new Ranges object that with modified or new metadata
columns or modified core components (start, end, width, seqnames, strand).
The use of `mutate()` means that a user no longer needs knowledge of the 
accessors of the Ranges object, as they can modify them in place.
The `filter()` function takes a Ranges object and logical expressions and
restricts Ranges object to where the logical expression evaluates to true.
The `summarise()` function takes a Ranges object and a set of name-value pairs
and aggregates the Ranges according to functions evaluated in the name-value pairs.
As `summarise()` is an aggregation it may break the structure of the of a Ranges
object, hence it returns a DataFrame object.
The `select()` function determines
which metadata columns are returned and the order they are returned in. The 
`arrange()` function sorts a Ranges object by named variables. 
The `group_by()` function creates an implicit grouping of Ranges object according
to variables in the Ranges object. This modifies the actions of `mutate()`,
`summarise()` and `filter()`, so they are performed on each partition created 
by the grouping. The `group_by()` operation acts as a replacement for 
the _GenomicRangesList_ class in the original _GenomicRanges_ API.

@@

The `reduce_ranges()` operation could be used to find coverage peaks 
accross chromosome 10. We can manually set a threshold to restrict genomic
regions to have a coverage score of greater than 8, and
then merge those regions and compute the maximum coverage over all the regions
that were reduced.

The combination of these verbs together encapsulate most operations that
can be performed in the original _IRanges_ and _GenomicRanges_ packages
without the user be exposed to new classes. In those packages to perform
the above operation requires users to understand what an run length encoding
object and views object means and perform additional manipulations to return
the result back to a _Ranges_.

### Arithimetic on Ranges

The API has an expressive algebra for performing arithmetic on Ranges via
the verbs `set_width()` and `stretch()`. As the names suggest `set_width()`
modifies the width of a Ranges object, while `stretch` extends the start and
end of a Ranges object. These can be chained with the anchoring functions
`anchor_start()`, `anchor_end()`, `anchor_center()`, `anchor_3p()` or 
`anchor_5p()`, which fix the coordinates of a Ranges object in place. 
Moreover, the `shift_` and `flank_` family of functions can be used to shift
all coordinates in a Ranges object or generate flanking regions from a Ranges 
object to the left, right, upstream or downstream of the input. Unlike, the
Bioconductor API, _plyranges_ makes it expclit via function calls whether to
take into account the strand information of a _Ranges_ object.

Returning to the Ranges object containing normalised coverage scores for
the methylation data, we can filter to find the coordinates of the peak 
containing maximum coverage score. We can then find 
a 5000 nt region centered around the maximum position by anchoring and 
modifying the the width.

### Overlapping Ranges

@@

These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges objects
as metadata. The intersect join takes the intersect of the start and end coordinates
of overlapping intervals of the query and subject Ranges 
(for GenomicRanges it also accounts for sequence name), when there is a overlap
the metadata corresponding to the query and subject Ranges are returned. 
Similarly, inner join takes the start and end coordinates of the query Ranges
that overlap the subject Ranges and returns metadata of the overlapping query
and subject Ranges. Finally, the left join performs a left outer join between
the query and subject Ranges, it returns all genomic intervals from the query
ranges, and returns missing values in metadata columns when there is no overlap.

The overlap innner join could be used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that contains
the max peak on chromosome 10 (visualised in figure \ref{fig:peak-viz}).

```{r peak_region}
peak_region <- chr10_scores %>%
  join_overlap_inner(., 
                     chr10_max_score_region %>% 
                       select(-score))

@@

A user may also restrict or group by overlaps with the `filter_by_overlaps()`,
`filter_by_non_overlaps()` and `group_by_overlaps()`. All overlap methods can
be modified with the `within` suffix (which changes the type of overlap 
from 'any' to 'within') or the `directed` suffix (which takes into account
the strand of a GenomicRanges object.).

For nearest neighbours, the `plyranges` API provides `join_nearest()`, 
`join_precede()`, and `join_follow()` functions. These functions are similar
to the overlapping functions, in that they return the query ranges that are
nearest (or precede or follow) the subject ranges and add metadata from
the subject ranges when the query is a nearest neighbour of the subject. Like
the overlap joins, these functions can modified with suffixes to find nearest
neighbours that are left, right, upstream or downstream of the subject.

The pairing operations, `pair_overlap()`, `pair_nearest()`, `pair_follow()`, 
and `pair_precede()` are similar to the join operation but instead of returning
a Ranges, they pair up the subject and query Ranges objects into a DataFrame,
alongside their metadata columns. This data structure is similar to the _Pairs_
data structure in the _S4Vectors_ package or the BED-PE file format. 
<!--- should also create convience methods for grouping by pairs and pairing up, 
unpairing, would also be useful for alignments too--->

- Redo an analysis in https://www.nature.com/articles/nature14248

in particular section: Trait-associated variants enrich in tissue-specific marks

@@

The _plyranges_ package is avaiable on the Bioconductor project website 
[https://bioconductor.org](https://bioconductor.org) or can be accessed
via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
We aim to continue developing the `plyranges` package and extend it for use
with more complex data structures  such as the 
_SummarizedExperiment_ class, which can be used for analysing transcriptomic
and variant data. As the `plyranges` interface encourages tidy data 
practices it integrates well with the principles of the grammar of graphics, 
we aim to use it for the visualisation of multimodal biological datasets.

>>>>

---
title: "plyranges: a grammar for manipulating genomics data" 
author:
  - name: Stuart Lee
    email: stuart.lee1@monash.edu
    affiliation: Monash University
  - name: Michael Lawrence
    email: michael.lawrence@gene.com
    affiliation: Genentech
  - name: Di Cook
    email: di.cook@monash.edu
    affiliation: Monash University
address:
  - code: Monash University
    address: Department of Econometrics and Business Statistics, Clayton, Victoria, Australia
  - code: Genentech
    address: Bioinformatics and Computational Biology, Genentech, Inc., South San Francisco, California, United States of America 
abstract: |
    The Bioconductor project has created many powerful abstractions for reasoning about genomics data, such as the _Ranges_ data structures for representing genomic intervals. By recognising that these data structures follow 'tidy' data principles we have created a grammar of genomic data manipulation that defines verbs for performing actions on and between genomic interval data. This grammar simplifies performing common genomic data analysis tasks via a consistent interface to existing Bioconductor infrastructure which results in creating human readable analysis pipelines. We have implemented this grammar as an Bioconductor/R package called plyranges.

bibliography: references.bib
output: 
  rticles::plos_article:
csl: plos.csl
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = "#>", 
                      fig.width = 5, fig.height = 3, fig.align = "center")
options(digits = 2)
library(ggbio)
library(ggplot2)
library(plyranges)

library(AnnotationHub)
q <- AnnotationHub()
bw_file <- q[["AH33458"]] 
```
# Introduction

Genomic data may be naturally represented as sets of pairs of integers 
corresponding to the start and end points of sequences. Further information
such as strandedness and chromosome name may be added to these sets to 
provide biological context. Because of the flexibility of this representation 
supplemental information such as measurements obtained from an experimental 
assay or annotations from a genome database can be joined to their relevant
sequences. In the Bioconductor/R packages `IRanges` and `GenomicRanges` these 
representations have been implemented as a suite of data structures called
_Ranges_ @Lawrence2013-wg; @Huber2015-ei. These data structures cover 
many common data types  encountered in bioinformatics analyses. For
example, a gene can be represented with its 
coordinates, along with its identifier and the identifiers of its exons; or
an RNA-seq experiment may be represented as sets of genes with a matching count 
column.

The Bioconductor infrastructure for computing with genomic ranges are highly 
efficient and powerful, however the application programming interface (API) for 
performing analysis tasks with _Ranges_ is complex due to its large number of
methods and classes. It also makes resulting scripts written 
difficult for a non-programmer to read and reason about. An alternative approach 
would be to implement a domain specific language (DSL) as a fluent interface 
built on top _Ranges_ [@Fowler2010-zd]. The goal of a fluent interface is to 
enable users to write  human-readable code via method chaining and consistent function returns. Fluent  interfaces fit naturally in the context of bioinformatics workflows because they 
enable writing succinct pipelines.

Several other DSLs have been implemented to reason about genomics data. Broadly, 
these are either implemented as query languages or as command line tools 
embedded in the Unix environment.  An example of the former is the 
Genome Query Language (GQL) and its distributed 
implementation GenAp which use an SQL-like syntax for fast retrieval of 
information from genomic databases and BAM files @Kozanitis2014-va; @Kozanitis2016-bm. 
Another  example is the Genometric Query Language (GMQL) which implements a 
relational algebra for combining big genomic datasets @Kaitoua2017-pw.  
The command line application 
BEDtools develops an extensive algebra for performing arithmetic between two or 
more sets of genomic regions  @Quinlan2010-gc. It also has a python interface 
which simplifies constructing scripts for performing analyses based on BEDTools 
@Dale2011-js. 
<!--- probably should also mention something about their cons, and more detail --->

The abstraction provided by the _Ranges_ data structures aligns with the concept
of tidy data @Wickham2014-jc. A tidy dataset is defined to be to a tabular
data structure that has observations as rows and columns as variables, and each
tidy dataset represents measurements from a single observational unit.
The tidy data pattern is useful because it allows 
us to see how the data relates to the design of an experiment
and the variables measured. Consequently, it makes both the modelling and 
manipulation of data more systematic. The _Ranges_ data structure follows 
this abstraction: it is a rectangular table corresponding to a single biological 
context. Each row contains a single observation and each column is a variable about
that observation.

The _plyranges_ API implements a domain specific language using the existing
_IRanges_ and _GenomicRanges_ packages in Bioconductor as a back-end. Consequently,
our API still has the speed and efficiency of the aforementioned packages but
with a more coherent interface. The API also
extends the grammar elements in _dplyr_ [@Wickham2017-dplyr] 
for performing genomic specific
manipulations such as finding overlapping regions or nearest neighbour regions
between many _Ranges_ . The _plyranges_ API is specifically designed to enable
fast interactive analysis of _Ranges_ objects but can also be used for scripting 
genomic data workflows.

@@

We have designed the API to be fluent. Every function call corresponds to an
action on a _Ranges_ object (they are named verbs) and where possible
functions have few arguments. Each verb is constructed
to enable a tab completion based workflow. Both of these aspects reduce the 
cognitive load on a new user since most manipulations can be performed with a 
vocabulary of several verbs, rather than having to memorise functions with many 
arguments that are nouns (as is required in the existing Bioconductor packages). 
This is also has the advantage of allowing users to write human readable code
because verbs describe what the code is doing rather than how its doing it.

Workflows can be composed by chaining verbs together via the forward pipe operator,`%>%` 
(exported from the R package _magrittr_ @R-magrittr). This is possible because every 
function call is an endomorphism: when the input is _Ranges_ object the output
will also be a _Ranges_ object. One advantage of endomorphism is that it 
does not  require any additional learning of classes beyond _Ranges_ and 
the _DataFrame_ classes. This strongly deviates from the design of 
the _Ranges_ Bioconductor packages, where many methods return a new class
upon return. The Bioconductor design enables efficient computing as users
are exposed to low-level features of its API which _plyranges_ abstracts
away. Method chaining via the pipe operator can also be difficult to debug,
as there multiple points of failure.

In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
use of non-standard evaluation in R via the _rlang_ package @R-rlang. Simply, this
means that computations are performed and evaluated in the context of the 
_Ranges_ objects, which emphasises the interactive nature of the API. This has
the disadvantage that programming with _plyranges_ becomes more difficult 
because a user needs to capture expressions inside function calls and then
unquote them.
<!--- need to make this clearer? --->

## Actions on Ranges

The _plyranges_ API exports the six core verbs from the _dplyr_ package
and modifies them for use with _Ranges_ objects. The
verb `mutate()` takes a Ranges object and a set of name-value pairs
and generates a new Ranges object that with modified or new metadata
columns or modified core components (start, end, width, seqnames, strand).
The use of `mutate()` means that a user no longer needs knowledge of the 
accessors of the Ranges object, as they can modify them in place.
The `filter()` function takes a Ranges object and logical expressions and
restricts Ranges object to where the logical expression evaluates to true.
The `summarise()` function takes a Ranges object and a set of name-value pairs
and aggregates the Ranges according to functions evaluated in the name-value pairs.
As `summarise()` is an aggregation it may break the structure of the of a Ranges
object, hence it returns a _DataFrame_ object.
The `select()` function determines
which metadata columns are returned and the order they are returned in. The 
`arrange()` function sorts a Ranges object by named variables. 
The `group_by()` function creates an implicit grouping of Ranges object according
to variables in the Ranges object. This modifies the actions of `mutate()`,
`summarise()` and `filter()`, so they are performed on each partition created 
by the grouping. The `group_by()` operation acts as a replacement for 
the _GenomicRangesList_ class in the original _GenomicRanges_ API.

@@

## Arithimetic on Ranges

The _plyranges_ API has an expressive algebra for performing arithmetic on Ranges via
the verbs `set_width()` and `stretch()`. As the names suggest `set_width()`
modifies the width of a Ranges object, while `stretch` extends the start and
end of a Ranges object. These can be chained with the anchoring functions
`anchor_start()`, `anchor_end()`, `anchor_center()`, `anchor_3p()` or 
`anchor_5p()`, which fix the coordinates of a Ranges object in place. 
Moreover, the `shift_` and `flank_` family of functions can be used to shift
all coordinates in a Ranges object or generate flanking regions from a Ranges 
object to the left, right, upstream or downstream of the input. Unlike, the
Bioconductor API, _plyranges_ makes it explicit via function calls whether to
take into account the strand information of a _Ranges_ object.

## Overlapping Ranges

@@

These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges objects
as metadata. The intersect join takes the intersect of the start and end coordinates
of overlapping intervals of the query and subject _Ranges_ 
(for _GenomicRanges_ it also accounts for sequence name), when there is a overlap
the metadata corresponding to the query and subject _Ranges_ are returned. 
Similarly, inner join takes the start and end coordinates of the query _Ranges_
that overlap the subject _Ranges_ and returns metadata of the overlapping query
and subject _Ranges_. Finally, the left join performs a left outer join between
the query and subject _Ranges_, it returns all genomic intervals from the query
ranges, and returns missing values in metadata columns when there is no overlap.

A user may also restrict or group by overlaps with the `filter_by_overlaps()`,
`filter_by_non_overlaps()` and `group_by_overlaps()`. All overlap methods can
be modified with the `within` suffix (which changes the type of overlap 
from 'any' to 'within') or the `directed` suffix (which takes into account
the strand of a _GenomicRanges_ object.).

For nearest neighbours, the _plyranges_ API provides `join_nearest()`, 
`join_precede()`, and `join_follow()` functions. These functions are similar
to the overlapping functions, in that they return the query ranges that are
nearest (or precede or follow) the subject ranges and add metadata from
the subject ranges when the query is a nearest neighbour of the subject. Like
the overlap joins, these functions can modified with suffixes to find nearest
neighbours that are left, right, upstream or downstream of the subject.

The pairing operations, `pair_overlap()`, `pair_nearest()`, `pair_follow()`, 
and `pair_precede()` are similar to the join operation but instead of returning
a _Ranges_, they pair up the subject and query _Ranges_ objects into a _DataFrame_,
alongside their metadata columns. This data structure is similar to the _Pairs_
data structure in the _S4Vectors_ @R-S4vectors package or the BED-PE file format.

The combination of verbs we have defined above encapsulate all operations that
can be performed in the original _IRanges_ and _GenomicRanges_ packages
without the user be exposed to new classes. In those packages to perform
most operations requires users to learn many classes and perform 
additional manipulations to return the results of their computation back to a
_Ranges_ object.

## Construction and Import/Output

The methods `as_granges()` and `as_iranges()`
for constructing _Ranges_ from tabular data structures, such as the _data.frame_ in
base R. These methods use non-standard evaluation so columns in a _data.frame_
can be modified before a _Ranges_ object is created. The API also has
convenience methods for annotating or extracting annotations from _GRanges_
objects with the `set_genome_info()` and `get_genome_info()` functions.

There is a consistent framework for reading and writing 
files from and to common genomic data formats, using the _rtracklayer_ package
as a back-end @Lawrence2009-nt. The methods are implemented in the `read_/write_` family of 
functions, currently _plyranges_ can read and write BAM, BED, BEDPE, 
narrowPeaks, GFF/GTF, WIG and BigWig files.

As an example, we use the Bioconductor package _AnnotationHub_ @R-ahub to search for 
BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project @Roadmap_Epigenomics_Consortium2015-pr.
We choose to focus on assays for primary T CD8+ memory cells from peripheral 
blood. We can then read the BigWig file  corresponding to the H3 lysine 27 
 trimethylation (H3K27Me3) 
methylation mark over chromosome 10.

First, we gather the BigWig file
and extract its annotation information and filter it to chromosome 10.

```{r load-bw, cache = TRUE}
library(plyranges)
chr10_ranges <- bw_file %>% 
  get_genome_info() %>%
  filter(seqnames == "chr10")
```

Then we read the BigWig file only extracting scores if they overlap chromosome
10. The annotation information from the file is automatically included (in this
case the hg19 genome build).

The `reduce_ranges()` operation is used to find coverage peaks 
across chromosome 10. We can manually set a threshold to restrict genomic
regions to have a coverage score of greater than 8, and
then merge nearby regions. The maximum coverage is computed over all the
coverage scores in regions that were reduced.

Returning to the _Ranges_ object containing normalised coverage scores for
the methylation data, we can filter to find the coordinates of the peak 
containing maximum coverage score. We can then find 
a 5000 nt region centered around the maximum position by anchoring and 
modifying the the width.

Finally, the overlap inner join could be used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that contains
the max peak on chromosome 10 (visualised in figure \ref{fig:peak-viz}).

```{r peak_region}
peak_region <- chr10_scores %>%
  join_overlap_inner(chr10_max_score_region %>% 
                       select(-score))

@@

@@

The _plyranges_ package is available on the Bioconductor project website 
[https://bioconductor.org](https://bioconductor.org) or can be accessed
via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
We aim to continue developing the _plyranges_ package and extend it for use
with more complex data structures  such as the 
_SummarizedExperiment_ class, which can be used for analysing transcriptomic
and variant data. As the _plyranges_ interface encourages tidy data 
practices it integrates well with the principles of the grammar of graphics, 
we aim to use it for the visualisation of multimodal biological datasets.
